### 2016-08-30 Projects/DelCo_2014_2015/Reports/Hose_Stream_Tests/TeX_Files/DelCo_Hose_Stream_Tests.tex
Edits to prepare final version of hose stream report

\vspace*{2\baselineskip}
July 2016}

@@

\large
\flushright{\bf National Institute of Standards and Technology Technical Note XXXX \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~XX, \pageref{LastPage} pages (July 2016) \\
% http://dx.doi.org/10.6028/NIST.TN.XXXX \\
CODEN: NTNOEF }

@@

% \chapter{List of Symbols}
% \begin{tabbing}
% \hspace{1.5in} \= \\
% ft \> foot \\
% gpm \> gallons per minute \\
% in \> inch \\
% m \> meter \\
% \end{tabbing}

@@

Fire suppression tactics using hose streams can affect ventilation in a structure and may impact the movement of smoke and heat through a structure. Seven series of experiments were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air movement within residential scale structures. The experiments studied four different hose streams: a straight stream, narrow fog stream, a wide fog stream from a combination nozzle, and a solid stream from a smooth bore nozzle. The streams were applied from a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Gas velocity was measured at different locations in the structure during the experiments. The wide fog stream caused the most air movement out of any of the tested streams, reaching a maximum velocity of 2.6~m/s (5.8~mph) and maximum air flow rate of approximately 4.2~m$^3$/s (9000~cfm), followed by the narrow fog stream, and then the straight stream and solid stream from the smooth bore nozzle, which both caused approximately the same amount of air movement through the structure. The data were consistent with a straight stream only causing air movement through the structure when it was applied in a moving pattern. Furthermore, there were no statistically significant differences in the average measured air velocity between the clockwise and counterclockwise nozzle movement patterns. It was determined that the type of hose stream and manner in which it is applied dictates the extent to which a stream impacts the ventilation of a structure. To better quantify the impact of such air flows on the fire environment, additional experiments need to be conducted with structure fires in a controlled environment.

@@

Four types of hose streams and four nozzle movement patterns were studied as part of the experiments described in this report. Three of the hose stream patterns studied were from a combination nozzle: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, experiments comparing the differences in impact on air flows in a structure from a combination nozzle straight stream and a solid stream from a smooth bore nozzle equipped with a 1~in tip were conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

\begin{figure}[!ht]
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/NF_example}
		\caption{Narrow fog stream from combination nozzle}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/WF_example}
		\caption{Wide fog stream from combination nozzle}
	\end{subfigure}
	\\~\\
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/SS_70}
		\caption{Straight stream from combination nozzle}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/SB_70}
		\caption{Solid stream from 1~in smooth bore nozzle}
	\end{subfigure}
	\caption[Hose stream patterns used during experiments.]{The four types of hose stream patterns used during water flow experiments.}
	\label{fig:hose_streams}
\end{figure}

@@

\subsubsection{Water Flow Rate}
Water flow rate was measured with a pressure and flow meter combination shown in Fig.~\ref{fig:flow_meter}. The meter consisted of a section of 63.5~mm (2.5~in) cast aluminum pipe with a 0 to 4.1~MPa (600~psi) pressure transducer and a paddlewheel type flow sensor with a range of 0 to 4800~lpm (1250~gpm). The pressure transducer and paddlewheel were both connected to the battery operated control box where the pressure transducer voltage was converted to a pressure and the paddlewheel pulse count was converted to a volumetric flow rate. The manufacturer reports a $\pm5$~\% calibration expanded uncertainty for the flow sensor and $\pm3$~\% for the pressure sensor~\cite{Akron:2009}. The pressure transducer was calibrated with a known analog pressure gauge. The flow meter was calibrated by capturing water over time and measuring that mass of water to determine the flow rate. The total expanded uncertainty was estimated at $\pm10$~\%.

@@

\section{Experimental Procedure}
\label{sec:exp_procedure}
Every experiment discussed in this report involved discharging water at a set flow rate from a combination nozzle in a straight, narrow fog, or wide fog stream pattern or from a smooth bore nozzle with a 1~in tip with a solid stream pattern. Specific doors were opened or closed before and/or during the tests to change the ventilation within the structure.

A total of seven experimental series, summarized in Table~\ref{table:test_matrix}, were conducted to study the impact of different hose stream and nozzle motion patterns on gas movement in a structure; three used a combination nozzle attached to a monitor to flow water at approximately 120~gpm, three used a combination nozzle attached directly to a 1.75~in handline to flow water at approximately 120~gpm (Fig.~\ref{fig:monitor+handline}), and one used both a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at approximately 180~gpm. Experiments that used a monitor to flow water always applied the stream in a fixed position and were focused on changing the location of water application in addition to the stream pattern. The experiments that used a combination nozzle attached directly to a 1.75~in handline to flow water focused on impact of the motion of the nozzle.

@@

\begin{figure}[!ht]
	\minipage{3in}
	\begin{center}
		\includegraphics[width=2.9in]{../Figures/Pictures/monitor}
	\end{center} 
	\endminipage \hfill
	\minipage{3in}
	\begin{center}
		\includegraphics[width=2.9in]{../Figures/Pictures/handline}
	\end{center}
	\endminipage \hfill
	\caption[Monitor and handline equipped with a combination nozzle.]{Monitor (left) and 1.75~in handline (right) equipped with a combination nozzle and used to flow water during experiments.}
	\label{fig:monitor+handline}
\end{figure}

@@

Two different flow path configurations were used for the test series conducted in the Two Story Structure. The main difference between the two flow paths was the position of the south side door on the first level. Tests~2 and 6 used a configuration with the door opened (Fig.~\ref{fig:flow_path_1}), while Tests~3, 4, and 7 used a configuration with the door closed (Fig.~\ref{fig:flow_path_2}). Tests~2 and 3 used a combination nozzle attached to a monitor to flow water; Tests~6 and 7 used a combination nozzle attached directly to a handline to flow water; and Test~4 used both a combination nozzle and a 1~in tip smooth bore nozzle attached to a monitor to flow water in a straight and solid stream, respectively.

@@

\subsection{Monitor Experiments}
\label{sec:monitor_procedure}
Three experimental series, Tests~1--3, used a monitor equipped with a combination nozzle to flow water at 120~gpm in a straight stream, narrow fog stream, and wide fog stream. Test~1 occurred in the Single Story Structure, and Tests~2 and 3 occurred in the Two Story Structure. A fourth series, Test~4, was conducted in the Two Story Structure and used a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at 180~gpm in a straight stream and solid stream, respectively. The monitor was always set to flow water from a fixed position. Thus, the monitor experiments primarily focused on the impact of different hose stream patterns on air movement in a structure and provided a baseline to compare the impact of the motion of the nozzle on the air movement through the structure.

@@

\subsubsection{Test~4}
Test~4 studied the differences in air movement caused by a solid stream from a smooth bore nozzle with a 1~in tip and straight stream from a combination nozzle attached to a monitor aimed at the same targets used during Tests~2 and 3. The stairwell door was opened for the entire experimental series. The test began by flowing water in a straight stream from a combination nozzle aimed at the near target. Then, the west double door on the second floor was opened. After 60 more seconds of flow, the nozzle was closed to stop water flow for 30 seconds. The nozzle was opened for 60 seconds and then closed for 30 seconds two additional times. Next, the monitor was aimed at the far target, and the procedure was repeated. This entire process was repeated for the solid stream from the smooth bore nozzle. Fig.~\ref{fig:test_4_pic} contains images of the straight stream from the combination nozzle and solid stream from the 1~in tip of the smooth bore nozzle during Test~4.

\begin{figure}[!ht]
	\minipage{2.15in}
	\begin{center}
		\includegraphics[width=2in]{../Figures/Pictures/SS_70}
	\end{center} 
	\endminipage
	\minipage{2.15in}
	\begin{center}
		\includegraphics[width=2in]{../Figures/Pictures/SB_70}
	\end{center}
	\endminipage
	\caption[Straight stream from combination nozzle and solid stream from smooth bore nozzle with 1~in tip during Test~4.]{Straight stream flowing from the combination nozzle (left) and solid stream flowing from the smooth bore nozzle (right) during Test~4.}
	\label{fig:test_4_pic}
\end{figure}
\FloatBarrier

\subsection{Handline Experiments}
\label{sec:handline_procedure}
Three experimental series were conducted using a 1.75~in handline with a combination nozzle to flow water in a straight stream, narrow fog stream, and wide fog stream. One series, Test~5, occurred in the Single Story Structure, and the remaining tests, Tests~6 and 7, occurred in the Two Story Structure. All three tests primarily focused on the impact of different hose stream and nozzle motion pattern selections on air movement in a structure.

\subsubsection{Test~5}
Test~5 utilized a 1.75~in handline equipped with a combination nozzle to flow water from Room C of the Single Story Structure into Room B (Fig.~\ref{fig:test_5_pic}). The layout of the structure was identical to the layout for Test~1, described above in Section~\ref{sec:monitor_procedure} and presented in Fig.~\ref{fig:east_setup}. The procedure for Test~5 began by flowing water in a straight stream aimed at the ceiling of Room B for 30 seconds in a fixed position. After 30 seconds, the nozzle was rotated in the clockwise direction for 30 seconds. Then, the nozzle was aimed at the south doorway of Room B, and water was applied for 30 seconds in a fixed position followed by 30 seconds in a clockwise pattern. The procedure was repeated using a narrow fog stream and a wide fog stream. This entire process was repeated for the streams being applied in the fixed and counterclockwise patterns.

@@

\subsubsection{Tests~6 \& 7}
Tests~6 and 7 followed identical procedures that involved using a 1.75~in handline equipped with a combination nozzle to flow water from the north side double doors of the Two Story Structure into the first floor room. The water stream was aimed at the south side wall of the first floor during the experiments. Test~6 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_1}, and Test~7 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_2}. Each test began by using a straight stream to flow water in a fixed, horizontal position. After 60 seconds of water flow, the stairwell door was opened. One minute later, the west double door on the north side of the second floor was opened, and water continued to flow for 60 seconds. Next, the water flow was stopped, the two doors were closed, and the procedure was repeated three additional times for the sweeping (back and forth, left-to-right), clockwise, and counterclockwise nozzle motion patterns. This entire process was repeated using narrow fog and wide fog streams. Fig.~\ref{fig:test_6_pic} contains an image of a straight stream pattern being applied during Test~6 with the flow path fully established.

@@

Looking at the table and figures, a number of trends can be observed. For example, there was air movement---an average gas velocity greater than 0.5~m/s (1.1~mph)---through the interior doorway while the flow path was fully established for every hose stream and target location combination except the straight stream in Test~1. Looking at Table~\ref{table:all_mon_vel_avgs}, the narrow fog and wide fog streams caused higher velocities through the interior doorway when the flow path was fully established compared to the straight stream. This trend can also be seen graphically in Fig.~\ref{fig:Test_2_BDP_A10_Avg_All} and Figs.~\ref{fig:Test_1_BDP_A6_Avg_All}~and~\ref{fig:Test_3_BDP_A10_Avg_All} in Appendix~\ref{chap:monitor_plots}. Furthermore, the wide fog stream generally produced equal or higher air velocities compared to the narrow fog stream. Finally, the Test~4 results listed in Table~\ref{table:all_mon_vel_avgs} and plotted in Fig.~\ref{fig:Test_4_BDP_A10_Avg_All} suggest that the straight stream from a combination nozzle and the solid stream from a smooth bore nozzle equipped with a 1~in tip produced approximately the same amount of air movement through the interior doorway.
\FloatBarrier

@@

With the exception of the narrow fog stream in the fixed position during Test~6, every hose stream and nozzle movement pattern combination using a narrow fog or wide fog stream during the handline experiments produced air movement---an average gas velocity greater than 0.5~m/s (1.1~mph)---through the interior doorway while the flow path was fully established. For the straight stream, however, air movement through the interior doorway only occurred when the stream was applied in a pattern that involved moving the handline. Similar to the monitor experiments, the average air velocity through the interior doorway was always greater in magnitude for the nozzle movement patterns using the narrow and wide fog streams than the same patterns using the straight stream. Another important result seen in Tables~\ref{table:east_hand_A6_avgs} and \ref{table:west_hand_A10_avgs} is that there are no statistically significant differences in the amount of air flow through the interior doorway when rotating the hoseline in the clockwise direction compared to rotating it in the counterclockwise direction for any of three hose streams tested. The similarities in air flow produced by the clockwise and counterclockwise nozzle movement patterns can also be seen graphically in Fig.~\ref{fig:Test_6_BDP_A10_Avg_All} and Figs.~\ref{fig:Test_5_BDP_A6_Avg_All} and \ref{fig:Test_7_BDP_A10_Avg_All} in Appendix~\ref{chap:handline_plots}.

@@

A total of seven experimental series were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air movement within residential scale structures.

@@

Additionally, a fourth experimental series using a monitor was conducted in which a straight stream from a combination nozzle was compared to a solid stream from a smooth bore nozzle with a 1~in tip aimed at a specified location for 60 seconds with a fully established flow path. The streams were compared for six different 60 second periods during which the flow path was fully established: three times each for the monitor aimed at two locations. The results from the four monitor test series suggest a wide fog stream impacts air movement within a structure the most out of any of the tested streams, followed by the narrow fog stream, and then the straight stream and solid stream from a smooth bore nozzle, which both caused approximately the same amount of air movement through the structure.

>>>>

\vspace*{2\baselineskip}
September 2016}

@@

\large
\flushright{\bf National Institute of Standards and Technology Technical Note XXXX \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~XX, \pageref{LastPage} pages (September 2016) \\
% http://dx.doi.org/10.6028/NIST.TN.XXXX \\
CODEN: NTNOEF }

@@

@@

Fire suppression tactics using hose streams can affect ventilation in a structure and may impact the movement of smoke and heat through a structure. Seven experimental series containing a total of 154 different configurations were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air movement within residential scale structures. The experiments studied four different hose streams: a straight stream, narrow fog stream, a wide fog stream from a combination nozzle, and a solid stream from a smooth bore nozzle. The streams were applied from a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Gas velocity was measured at different locations in the structure during the experiments. The wide fog stream caused the most air movement out of any of the tested streams, reaching a maximum velocity of 2.6~m/s (5.8~mph) and maximum air flow rate of approximately 4.2~m$^3$/s (9000~cfm), followed by the narrow fog stream, and then the straight stream and solid stream from the smooth bore nozzle, which both caused approximately the same amount of air movement through the structure. The data were consistent with a straight stream only causing air movement through the structure when it was applied in a moving pattern. Furthermore, there were no statistically significant differences in the average measured air velocity between the clockwise and counterclockwise nozzle movement patterns. It was determined that the type of hose stream and manner in which it is applied dictates the extent to which a stream impacts the ventilation of a structure. To better quantify the impact of such air flows on the fire environment, additional experiments need to be conducted with structure fires in a controlled environment.

@@

Four types of hose streams and four nozzle movement patterns were studied under 154 different configurations during the seven test series described in this report. Three of the hose stream patterns studied were from a combination nozzle: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, experiments comparing the differences in impact on air flows in a structure from a combination nozzle straight stream and a solid stream from a smooth bore nozzle equipped with a 1~in. tip were conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

\begin{figure}[!ht]
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/NF_example}
		\caption{Narrow fog stream from combination nozzle}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/WF_example}
		\caption{Wide fog stream from combination nozzle}
	\end{subfigure}
	\\~\\
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/SS_70}
		\caption{Straight stream from combination nozzle}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\columnwidth}
		\centering
		\includegraphics[width=0.75\columnwidth]{../Figures/Pictures/SB_70}
		\caption{Solid stream from 1~in. smooth bore nozzle}
	\end{subfigure}
	\caption[Hose stream patterns used during experiments.]{The four types of hose stream patterns used during water flow experiments.}
	\label{fig:hose_streams}
\end{figure}

@@

\subsubsection{Water Flow Rate}
Water flow rate was measured with a pressure and flow meter combination shown in Fig.~\ref{fig:flow_meter}. The meter consisted of a section of 63.5~mm (2.5~in) cast aluminum pipe with a 0 to 4.1~MPa (600~psi) pressure transducer and a paddlewheel type flow sensor with a range of 0 to 4800~lpm (1250~gpm). The pressure transducer and paddlewheel were both connected to the battery operated control box where the pressure transducer voltage was converted to a pressure and the paddlewheel pulse count was converted to a volumetric flow rate. The manufacturer reports a $\pm5$~\% calibration expanded uncertainty for the flow sensor and $\pm3$~\% for the pressure sensor~\cite{Akron:2009}. The pressure transducer was calibrated with a known analog pressure gauge. The flow meter was calibrated by capturing water over time and measuring that mass of water to determine the flow rate. The total expanded uncertainty was estimated to be $\pm10$~\%.

@@

\section{Experimental Procedure}
\label{sec:exp_procedure}
Every experimental test series discussed in this report involved discharging water at a set flow rate from a combination nozzle in a straight, narrow fog, or wide fog stream pattern or from a smooth bore nozzle with a 1~in. tip in a solid stream pattern. Specific doors were opened or closed before and/or during the tests to change the ventilation within the structure. Each time a hose stream pattern, nozzle movement pattern, target location, or ventilation pattern was changed during a test series, a new test configuration was considered to be in place. A total of 154 configurations were used throughout the seven experimental series.

The seven experimental test series, summarized in Table~\ref{table:test_matrix}, were conducted to study the impact of different hose stream and nozzle motion patterns on gas movement in a structure; three used a combination nozzle attached to a monitor to flow water at approximately 120~gpm, three used a combination nozzle attached directly to a 1.75~in. handline to flow water at approximately 120~gpm (Fig.~\ref{fig:monitor+handline}), and one used both a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at approximately 180~gpm. Experiments that used a monitor to flow water always applied the stream in a fixed position and were focused on changing the location of water application in addition to the stream pattern. The experiments that used a combination nozzle attached directly to a 1.75~in. handline to flow water focused on impact of the motion of the nozzle.

@@

\begin{figure}[!ht]
	\minipage{3in}
	\begin{center}
		\includegraphics[width=2.9in]{../Figures/Pictures/monitor}
	\end{center} 
	\endminipage \hfill
	\minipage{3in}
	\begin{center}
		\includegraphics[width=2.9in]{../Figures/Pictures/handline}
	\end{center}
	\endminipage \hfill
	\caption[Monitor and handline equipped with a combination nozzle.]{Monitor (left) and 1.75~in. handline (right) equipped with a combination nozzle and used to flow water during experiments.}
	\label{fig:monitor+handline}
\end{figure}

@@

Two different flow path configurations were used for the test series conducted in the Two Story Structure. The main difference between the two flow paths was the position of the south side door on the first level. Tests~2 and 6 used a configuration with the door opened (Fig.~\ref{fig:flow_path_1}), while Tests~3, 4, and 7 used a configuration with the door closed (Fig.~\ref{fig:flow_path_2}). Tests~2 and 3 used a combination nozzle attached to a monitor to flow water; Tests~6 and 7 used a combination nozzle attached directly to a handline to flow water; and Test~4 used both a combination nozzle and a 1~in. tip smooth bore nozzle attached to a monitor to flow water in a straight and solid stream, respectively.

@@

\subsection{Monitor Experiments}
\label{sec:monitor_procedure}
Three experimental test series, Tests~1--3, used a monitor equipped with a combination nozzle to flow water at 120~gpm in a straight stream, narrow fog stream, and wide fog stream. Test~1 occurred in the Single Story Structure, and Tests~2 and 3 occurred in the Two Story Structure. A fourth test series, Test~4, was conducted in the Two Story Structure and used a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at 180~gpm in a straight stream and solid stream, respectively. The monitor was always set to flow water from a fixed position. Thus, the monitor experiments primarily focused on the impact of different hose stream patterns on air movement in a structure and provided a baseline to compare the impact of the motion of the nozzle on the air movement through the structure.

@@

\subsubsection{Test~4}
Test~4 studied the differences in air movement caused by a solid stream from a smooth bore nozzle with a 1~in. tip and straight stream from a combination nozzle attached to a monitor aimed at the same targets used during Tests~2 and 3. The stairwell door was opened for the entire experimental series. The test began by flowing water in a straight stream from a combination nozzle aimed at the near target. Then, the west double door on the second floor was opened. After 60 more seconds of flow, the nozzle was closed to stop water flow for 30 seconds. The nozzle was opened for 60 seconds and then closed for 30 seconds two additional times. Next, the monitor was aimed at the far target, and the procedure was repeated. This entire process was repeated for the solid stream from the smooth bore nozzle. Fig.~\ref{fig:test_4_pic} contains images of the straight stream from the combination nozzle and solid stream from the 1~in. tip of the smooth bore nozzle during Test~4.

\begin{figure}[!ht]
	\minipage{2.15in}
	\begin{center}
		\includegraphics[width=2in]{../Figures/Pictures/SS_70}
	\end{center} 
	\endminipage
	\minipage{2.15in}
	\begin{center}
		\includegraphics[width=2in]{../Figures/Pictures/SB_70}
	\end{center}
	\endminipage
	\caption[Straight stream from combination nozzle and solid stream from smooth bore nozzle with 1~in. tip during Test~4.]{Straight stream flowing from the combination nozzle (left) and solid stream flowing from the smooth bore nozzle (right) during Test~4.}
	\label{fig:test_4_pic}
\end{figure}
\FloatBarrier

\subsection{Handline Experiments}
\label{sec:handline_procedure}
Three experimental test series were conducted using a 1.75~in. handline with a combination nozzle to flow water in a straight stream, narrow fog stream, and wide fog stream. One series, Test~5, occurred in the Single Story Structure, and the remaining tests, Tests~6 and 7, occurred in the Two Story Structure. All three tests primarily focused on the impact of different hose stream and nozzle motion pattern selections on air movement in a structure.

\subsubsection{Test~5}
Test~5 utilized a 1.75~in. handline equipped with a combination nozzle to flow water from Room C of the Single Story Structure into Room B (Fig.~\ref{fig:test_5_pic}). The layout of the structure was identical to the layout for Test~1, described above in Section~\ref{sec:monitor_procedure} and presented in Fig.~\ref{fig:east_setup}. The procedure for Test~5 began by flowing water in a straight stream aimed at the ceiling of Room B for 30 seconds in a fixed position. After 30 seconds, the nozzle was rotated in the clockwise direction for 30 seconds. Then, the nozzle was aimed at the south doorway of Room B, and water was applied for 30 seconds in a fixed position followed by 30 seconds in a clockwise pattern. The procedure was repeated using a narrow fog stream and a wide fog stream. This entire process was repeated for the streams being applied in the fixed and counterclockwise patterns. The average rotation rate for configurations that involved nozzle rotation during Test~5 was 2.1 rotations per second with a maximum rate of 2.3 rotations per second and minimum of 1.9 rotations per second.

@@

\subsubsection{Tests~6 \& 7}
Tests~6 and 7 followed identical procedures that involved using a 1.75~in. handline equipped with a combination nozzle to flow water from the north side double doors of the Two Story Structure into the first floor room. The water stream was aimed at the south side wall of the first floor during the experiments. Test~6 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_1}, and Test~7 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_2}. Each test began by using a straight stream to flow water in a fixed, horizontal position. After 60 seconds of water flow, the stairwell door was opened. One minute later, the west double door on the north side of the second floor was opened, and water continued to flow for 60 seconds. Next, the water flow was stopped, the two doors were closed, and the procedure was repeated three additional times for the sweeping (back and forth, left-to-right), clockwise, and counterclockwise nozzle motion patterns. This entire process was repeated using narrow fog and wide fog streams. Fig.~\ref{fig:test_6_pic} contains an image of a straight stream pattern being applied during Test~6 with the flow path fully established. During configurations involving nozzle rotation with a wide fog stream or narrow fog stream, the average rotation rate was 1.3 rotations per second with a maximum of 1.4 rotations per second and a minimum of 1.2 rotations per second. The rotation rates for the clockwise and counterclockwise rotation of the straight stream were 0.8 rotations per second and 0.7 rotations per second, respectively, for Test~6, and 0.4 rotations per second for both directions during Test~7.

@@

Looking at the table and figures, a number of trends can be observed. For example, there was air movement---an average gas velocity greater than 0.5~m/s (1.1~mph)---through the interior doorway while the flow path was fully established for every hose stream and target location combination except the straight stream in Test~1. Looking at Table~\ref{table:all_mon_vel_avgs}, the narrow fog and wide fog streams caused higher velocities through the interior doorway when the flow path was fully established compared to the straight stream. This trend can also be seen graphically in Fig.~\ref{fig:Test_2_BDP_A10_Avg_All} and Figs.~\ref{fig:Test_1_BDP_A6_Avg_All}~and~\ref{fig:Test_3_BDP_A10_Avg_All} in Appendix~\ref{chap:monitor_plots}. Furthermore, the wide fog stream generally produced equal or higher air velocities compared to the narrow fog stream. Finally, the Test~4 results listed in Table~\ref{table:all_mon_vel_avgs} and plotted in Fig.~\ref{fig:Test_4_BDP_A10_Avg_All} suggest that the straight stream from a combination nozzle and the solid stream from a smooth bore nozzle equipped with a 1~in. tip produced approximately the same amount of air movement through the interior doorway.
\FloatBarrier

@@

With the exception of the narrow fog stream in the fixed position during Test~6, every hose stream and nozzle movement pattern combination using a narrow fog or wide fog stream during the handline experiments produced air movement---an average gas velocity greater than 0.5~m/s (1.1~mph)---through the interior doorway while the flow path was fully established. For the straight stream, however, air movement through the interior doorway only occurred when the stream was applied in a pattern that involved moving the handline. Similar to the monitor experiments, the average air velocity through the interior doorway was always greater in magnitude for the nozzle movement patterns using the narrow and wide fog streams than the same patterns using the straight stream. Another important result seen in Tables~\ref{table:east_hand_A6_avgs} and \ref{table:west_hand_A10_avgs} is that there are no statistically significant differences (differences greater than the error associated with the bi-directional probes) in the amount of air flow through the interior doorway when rotating the hoseline in the clockwise direction compared to rotating it in the counterclockwise direction for any of three hose streams tested. The similarities in air flow produced by the clockwise and counterclockwise nozzle movement patterns can also be seen graphically in Fig.~\ref{fig:Test_6_BDP_A10_Avg_All} and Figs.~\ref{fig:Test_5_BDP_A6_Avg_All} and \ref{fig:Test_7_BDP_A10_Avg_All} in Appendix~\ref{chap:handline_plots}.

@@

Seven experimental test series containing a total of 154 different configurations were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air movement within residential scale structures.

@@

Additionally, a fourth experimental series using a monitor was conducted in which a straight stream from a combination nozzle was compared to a solid stream from a smooth bore nozzle with a 1~in. tip aimed at a specified location for 60 seconds with a fully established flow path. The streams were compared for six different 60 second periods during which the flow path was fully established: three times each for the monitor aimed at two locations. The results from the four monitor test series suggest a wide fog stream impacts air movement within a structure the most out of any of the tested streams, followed by the narrow fog stream, and then the straight stream and solid stream from a smooth bore nozzle, which both caused approximately the same amount of air movement through the structure.
